---
title: "Tables in Emails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables in Emails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

`mc_md_render()` converts a single markdown file to HTML. But what if you need
an R-generated table in the middle of your email? You can't put R code in a
plain `.md` file.

## mc_compose()

`mc_compose()` takes any mix of markdown files, HTML strings, and kable objects
and stitches them into one email body.

```{r compose-basic, eval = FALSE}
library(mc)

df <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000)
)

body <- mc_compose(
  "communications/intro.md",
  knitr::kable(df, format = "html"),
  "<p>Let me know if the numbers look right.</p>"
)

mc_send(html = body,
        to = "brandon@example.com",
        subject = "Planting site summary")
```

Each argument is resolved in order:

- **`.md` file** — header above `---` is stripped, body converted to HTML
- **kable/kableExtra object** — converted to HTML via `as.character()`
- **HTML string** — passed through as-is

All `<table>`, `<th>`, and `<td>` elements get border and padding styles
injected automatically for Gmail compatibility.

## Simple table with knitr::kable

The most reliable option. Clean HTML, minimal styling, works everywhere.

```{r kable-example}
df <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Status = c("Confirmed", "Pending", "Pending")
)

knitr::kable(df, format = "html")
```

`mc_compose()` adds border and padding inline styles to every cell.

## Styled table with kableExtra

For colors, bold headers, or highlighted rows, use `kableExtra::row_spec()`
and `kableExtra::column_spec()`. These add **inline styles** that survive
Gmail's CSS stripping.

```{r kableextra-example}
library(kableExtra)

kbl(df, format = "html") |>
  kable_styling(full_width = FALSE) |>
  row_spec(0, bold = TRUE, background = "#f5f5f5") |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, color = "#2c7bb6")
```

**Avoid** `bootstrap_options` like `"striped"` or `"hover"` — these rely on
CSS classes in a `<style>` block that Gmail strips. Use `row_spec()` with
explicit `background` colors instead:

```{r striping-manual, eval = FALSE}
# Manual striping that works in Gmail
kbl(df, format = "html") |>
  kable_styling(full_width = FALSE) |>
  row_spec(0, bold = TRUE, background = "#f5f5f5") |>
  row_spec(seq(2, nrow(df), 2), background = "#f9f9f9")
```

## Scrollable tables

Large tables can be wrapped in a scrollable `<div>` and Gmail preserves the
`overflow` CSS. Use `mc_scroll()` to wrap any table HTML.

**Wide table** (horizontal scroll):

```{r scroll-wide, eval = FALSE}
wide_df <- data.frame(...)  # many columns

body <- mc_compose(
  "<p>Wide table with horizontal scroll:</p>",
  mc_scroll(knitr::kable(wide_df, format = "html"), direction = "wide")
)
```

**Long table** (vertical scroll, capped height):

```{r scroll-long, eval = FALSE}
long_df <- data.frame(...)  # many rows

body <- mc_compose(
  "<p>Long table with vertical scroll:</p>",
  mc_scroll(knitr::kable(long_df, format = "html"), direction = "long")
)
```

**Both directions:**

```{r scroll-both, eval = FALSE}
big_df <- data.frame(...)  # many rows and columns

body <- mc_compose(
  "<p>Big table scrolling both ways:</p>",
  mc_scroll(knitr::kable(big_df, format = "html"), direction = "both")
)
```

The default `max_height` for vertical scrolling is `"400px"`. Adjust with
`mc_scroll(..., max_height = "600px")`.

## Full workflow

```{r full-workflow, eval = FALSE}
library(mc)

# Authenticate once per session
mc_auth()

# Build the email
sites <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Season = c("Fall 2026", "Fall 2026", "Fall 2026")
)

body <- mc_compose(
  "<p>Hi Brandon,</p>
   <p>Here's the current planting plan:</p>",

  kableExtra::kbl(sites, format = "html") |>
    kableExtra::kable_styling(full_width = FALSE) |>
    kableExtra::row_spec(0, bold = TRUE, background = "#f5f5f5"),

  "<p>Total: 10,000 plugs across three sites.</p>
   <p>Does this match what the nursery has available?</p>"
)

# Draft first, send when ready
mc_send(html = body,
        to = "brandon@example.com",
        subject = "2026 planting plan — cottonwood plugs")
```
