---
title: "Tables in Emails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables in Emails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mc)
```

## The problem

`mc_md_render()` converts a single markdown file to HTML. But what if you need
an R-generated table in the middle of your email? You can't put R code in a
plain `.md` file.

## mc_compose()

`mc_compose()` takes any mix of markdown files, HTML strings, and kable objects
and stitches them into one email body.

```{r compose-basic, eval = FALSE}
library(mc)

df <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000)
)

body <- mc_compose(
  "communications/intro.md",
  knitr::kable(df, format = "html"),
  "<p>Let me know if the numbers look right.</p>"
)

mc_send(html = body,
        to = "brandon@example.com",
        subject = "Planting site summary")
```

Each argument is resolved in order:

- **`.md` file** — header above `---` is stripped, body converted to HTML
- **kable/kableExtra object** — converted to HTML via `as.character()`
- **HTML string** — passed through as-is

All `<table>`, `<th>`, and `<td>` elements get border and padding styles
injected automatically for Gmail compatibility.

## Simple table with knitr::kable

The most reliable option. Clean HTML, minimal styling, works everywhere.

```{r kable-simple}
df <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Status = c("Confirmed", "Pending", "Pending")
)

knitr::kable(df, format = "html")
```

`mc_compose()` adds border and padding inline styles to every cell.

## Styled table with kableExtra

For colors, bold headers, or highlighted rows, use `kableExtra::row_spec()`
and `kableExtra::column_spec()`. These add **inline styles** that survive
Gmail's CSS stripping.

```{r kableextra-styled}
library(kableExtra)

kbl(df, format = "html") |>
  kable_styling(full_width = FALSE) |>
  row_spec(0, bold = TRUE, background = "#f5f5f5") |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, color = "#2c7bb6")
```

**Avoid** `bootstrap_options` like `"striped"` or `"hover"` — these rely on
CSS classes in a `<style>` block that Gmail strips. Use `row_spec()` with
explicit `background` colors instead:

```{r striping-manual}
kbl(df, format = "html") |>
  kable_styling(full_width = FALSE) |>
  row_spec(0, bold = TRUE, background = "#f5f5f5") |>
  row_spec(seq(2, nrow(df), 2), background = "#f9f9f9")
```

## Scrollable tables

Large tables can be wrapped in a scrollable `<div>` using `mc_scroll()`.
Gmail preserves the `overflow` CSS so these scroll in the email client.

### Wide table (horizontal scroll)

```{r scroll-wide-live, results = "asis"}
wide_df <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Season = c("Fall 2026", "Fall 2026", "Fall 2026"),
  Latitude = c(53.92, 55.34, 54.38),
  Longitude = c(-124.02, -122.97, -127.17),
  BEC_Zone = c("SBSdw3", "SBSmk1", "SBSdk"),
  Elevation_m = c(680, 700, 520),
  Stock = c("Local", "Transfer", "Transfer"),
  Nursery = c("PRT Vernon", "PRT Vernon", "PRT Vernon"),
  Crew = c("Alpine Reforestation", "TBD", "TBD"),
  Access = c("FSR 200", "FSR 440", "Morice West"),
  Prep = c("Complete", "Scheduled", "Not started"),
  Soil = c("Silty clay", "Sandy loam", "Clay loam"),
  Moisture = c("Subhygric", "Mesic", "Hygric"),
  Permit = c("Approved", "In review", "Not submitted")
)

cat(mc_scroll(knitr::kable(wide_df, format = "html"), direction = "wide"))
```

### Long table (vertical scroll)

```{r scroll-long-live, results = "asis"}
set.seed(42)
sites <- c("Nechako", "Mackenzie", "Skeena", "Fraser", "Bulkley",
           "Stellako", "Endako", "Stuart", "Salmon", "Bowron")
long_df <- data.frame(
  Row = 1:50,
  Site = rep(sites, 5),
  Plugs = sample(100:500, 50, replace = TRUE),
  Survival_Pct = round(runif(50, 60, 98), 1),
  Height_cm = round(runif(50, 15, 85), 0)
)

cat(mc_scroll(knitr::kable(long_df, format = "html"), direction = "long"))
```

### Both directions

```{r scroll-both-live, results = "asis"}
set.seed(42)
both_df <- data.frame(
  Row = 1:30,
  Site = rep(sites, 3),
  Plugs = sample(100:500, 30, replace = TRUE),
  Survival = round(runif(30, 60, 98), 1),
  Height_cm = round(runif(30, 15, 85), 0),
  BEC_Zone = rep(c("SBSdw3", "SBSmk1", "SBSdk", "SBSwk1", "SBSmc2",
                   "SBSdw1", "SBSwk3", "SBSdk", "SBSmw", "SBSdw2"), 3),
  Elevation = sample(400:900, 30, replace = TRUE),
  Lat = round(runif(30, 53, 56), 4),
  Lon = round(runif(30, -128, -122), 4),
  Source = rep(c("Local", "Transfer"), 15),
  Soil = rep(c("Silty clay", "Sandy loam", "Clay loam", "Silt loam", "Gravel"), 6),
  Moisture = rep(c("Subhygric", "Mesic", "Hygric", "Subxeric", "Mesic"), 6),
  Fish = rep(c("Yes", "Yes", "No", "Yes", "No"), 6),
  Permit = rep(c("Approved", "In review", "Not submitted", "Approved", "Pending"), 6)
)

cat(mc_scroll(knitr::kable(both_df, format = "html"), direction = "both"))
```

The default `max_height` for vertical scrolling is `"400px"`. Adjust with
`mc_scroll(..., max_height = "600px")`.

## Full workflow

```{r full-workflow, eval = FALSE}
library(mc)

sites <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Season = c("Fall 2026", "Fall 2026", "Fall 2026")
)

body <- mc_compose(
  "<p>Hi Brandon,</p>
   <p>Here's the current planting plan:</p>",

  kableExtra::kbl(sites, format = "html") |>
    kableExtra::kable_styling(full_width = FALSE) |>
    kableExtra::row_spec(0, bold = TRUE, background = "#f5f5f5"),

  "<p>Total: 10,000 plugs across three sites.</p>
   <p>Does this match what the nursery has available?</p>"
)

mc_send(html = body,
        to = "brandon@example.com",
        subject = "2026 planting plan — cottonwood plugs")
```

## Scheduled send

Use `send_at` to send an email later. This works with `mc_compose()` and
`mc_scroll()` — build the body now, send it when you want.

```{r scheduled-send, eval = FALSE}
library(mc)

sites <- data.frame(
  Site = c("Nechako", "Mackenzie", "Skeena"),
  Plugs = c(4000, 3000, 3000),
  Season = c("Fall 2026", "Fall 2026", "Fall 2026")
)

body <- mc_compose(
  "<p>Hi Brandon,</p>
   <p>Here's the planting plan with a scrollable table:</p>",
  mc_scroll(
    kableExtra::kbl(sites, format = "html") |>
      kableExtra::kable_styling(full_width = FALSE) |>
      kableExtra::row_spec(0, bold = TRUE, background = "#f5f5f5"),
    direction = "both"
  ),
  "<p>Let me know if the numbers look right.</p>"
)

# Send in 30 minutes
proc <- mc_send(html = body,
                to = "brandon@example.com",
                subject = "2026 planting plan",
                send_at = 30)

# Or at a specific time
mc_send(html = body,
        to = "brandon@example.com",
        subject = "2026 planting plan",
        send_at = as.POSIXct("2026-02-24 09:00:00"))

# Check or cancel
proc$is_alive()
proc$kill()
```

On macOS, `caffeinate` prevents the machine from sleeping until the email
sends. The laptop lid can be closed as long as power is connected. If the
machine does sleep through the send window, a 5-minute grace period applies
— past that the send is skipped to prevent stale emails firing.
